
---
title: "Population Projections Example"
output: html_document
date: "`r format(Sys.Date())`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



This R script allows the user to project future demand based on population projections sourced from ONS. 
The population projections are from the excel file called table2 and the data to be projected needs to be 
in the correct format, detailed in the excel file attached. 
The projection data has seperate tabs for different genders, along with a combined dataset for 
both genders. The below script projects the combined dataset (Persons sheet). 

``` {r table 1, include = FALSE, results = "markup"}

library(qcc)
library(tidyverse)
library(readxl)
library(openxlsx)
library(lubridate)
library(ggplot2)
library(scales)
library(dygraphs)
library(rlang)
library(writexl)
library(DBI)
library(odbc)
library(scales)

```



```{r table 2}

Population <- read_xlsx(path = "table2.xlsx", sheet = "Persons")

head(Population, 10)

```

The population data contains a Code column which will not be required for any analysis and 
also contains data for 'All Ages' which needs to be removed to ensure there's no duplication of 
analysis. Additionally, the data is in a format where the years are column headings. To make 
analysis easier we need to pivot the data longer. 

As the population data contains data from 2018 to 2043 we need to remove years 2018 to 2021.

```{r table 3}
Population <-  Population %>% filter(`AGE GROUP` != "All ages")

Population <- Population %>% subset(select = -c(CODE))

Population <- Population %>% pivot_longer(`2018`:`2043`, names_to = "Year", values_to = "Population")

Population <- Population %>% filter(Year != 2018)

Population <- Population %>% filter(Year != 2019)

Population <- Population %>% filter(Year != 2020)

Population <- Population %>% filter(Year != 2021)


```

The population data contains information for all regions within England, however this 
analysis was focussed on Nottinghamshire. Please change the filter accordingly. 

```{r table 4}

Population <- Population %>% filter(AREA %in% c("Ashfield", "Bassetlaw", "Broxtowe", "Gedling", "Mansfield", "Newark and Sherwood", "Nottingham", "Rushcliffe"))

```

We now need to read in the example dataset to project forward using the population data. 
This has been provided as a template with example data. Please change the values accordingly

``` {r table 5}

Values <- read_xlsx(path = "Values.xlsx")

head(Values, 10)

```

In order to project the data we need to merge the population data with the projection data. 
To do this we need to make a unique indexing column to perform the left join which is
a combination of the area and age group. This is done using the paste0 function. 

``` {r table 6}

Population <-  Population %>% mutate(Index = paste0(AREA, `AGE GROUP`))

Values <- Values %>% mutate(Index = paste0(Area, `Age Group`))

Projection_Data <- left_join(Population, Values, by = "Index")


```

The left join will create duplicate columns which need to be removed. Additionally,
tidying the data will make projections easier. 

``` {r table 7}

Projection_Data <- Projection_Data %>% subset(select = c(Area, `Age Group`, Year, Population, Index, Values))

```

We can now calculate the projections. Due to the high number of unique groups within the dataset, 
calculating the projections one-by-one would be time consuming. Therefore we create a loop
which will run over the index column to calculate the projections for each of these 
unique groups. In order to create a new dataset with the calculated projections in we need
to create a new, empty dataframe. 

``` {r table 8}

df_1 <- data.frame()

for (i in unique(Projection_Data$Index)) {
  Projection_Data_2 <- Projection_Data %>% filter(Index == i) %>% mutate(Pop_Growth = Population/head(Population, n = 1)) %>%
    mutate(Value_Growth = Values*Pop_Growth)
  df_1 <- rbind(df_1, Projection_Data_2)
}


```


Now that the data has been cleaned and formatted we can visualise the projections in 
the plots below. 

``` {r table 10, echo = FALSE}

df_1 %>% ggplot(aes(x = Year, y = Value_Growth)) +
  geom_col() +
  theme_bw() +
  scale_x_discrete(breaks = c("2025", "2030", "2035", "2040")) +
  labs(y = "Growth", title = "Total Projected Growth to 2043")

df_1 %>% ggplot(aes(x = Year, y = Value_Growth)) +
  geom_col() +
  theme_bw() +
  facet_wrap(~`Age Group`) +
  scale_x_discrete(breaks = c("2030", "2040")) +
  labs(y = "Growth", title = "Total Projected Growth to 2043", subtitle = "Split by Age Group")



```


As the projections have calculated for the individual areas we can create a loop 
to automatically plot the individual graphs

``` {r table 11}

for(i in unique(df_1$Area)){
  title_label = paste0("Projected Increase by Age Group in ", i)
  
  print(df_1 %>% filter(Area == i) %>% ggplot(aes(x = Year, y = Value_Growth)) +
  geom_col() +
  facet_wrap(~`Age Group`, nrow = 5) +
    theme_bw() +
    labs(title = title_label, y = "Growth") +
    scale_x_discrete(breaks = c("2025", "2030", "2035", "2040")))}

```
